---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024`- `Fall 2021`)/`Fall 2021` * 100)
```

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2025`- `Fall 2024`)/`Fall 2024` * 100)
```


### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?

Most of the percent changes increased by marginal amounts, less than 5%. However, two values increased by more than 5%, shuttle bus and student activities. Shuttle bus fees increased by 9% which sticks out but may be chalked up to increasing wages inline with the economy. Further investigation would still be needed. Student activities increased by 14.6% which is the largest jump in the table. This could be the story because why would it increases so much? Are students getting more while paying more? Have student activities become more expensive to offer for the school? 



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.


```{r}
comp <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/class_composition.csv")

comp |>
  summarize(
    mean_first = mean(first_year),
    mean_soph = mean(sophomore),
    mean_junior = mean(junior),
    mean_senior = mean(senior),
    mean_grad = mean(grad)
  )
```

