---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Danny Martella"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
res <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- res |>
  summarise(
    mean_comp = mean(compliance_score),
  sd_comp = sd(compliance_score) ,
  min_comp = min(compliance_score),
max_comp = max(compliance_score)
  )
  
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
res |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth =1)+
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) - sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) + sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
  labs(
    title = "Montgomery County Restaurant Health Inspection Compliance Scores",
    x = "Compliance Score" , 
    y = "Count"
  )
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**

Based on the histogram and the compliance stats, we can see that most of the restaurants received high compliance scores. The smaller standard deviation tells us a vast majority of the data falls close to the mean, within one standard deviation. I think the fact that so many restaurants got a score of 100 is newsworthy. Even though we expect restaurants to comply with basic health inspection standards, it is good to know that most of the restaurants people are going to are clean and safe. I also think its important to be aware of the places that scored lower. If I were to write a story, my next steps would be to visit the lower scoring restaurants to see how dirty they look, then reach out to ownership and see if they will try to make changes to their practices to improve their score. 

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata |>
  summarise(
    avg_bus = mean(bus),
    sd_bus = sd(bus),
    avg_rail = mean(rail),
    sd_rail = sd(rail)
  )
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata |>
  mutate(
    bus_zscore = (total - mean(bus)) / sd(bus) , 
    rail_zscore = (total - mean(rail)) / sd(rail)
  )
  
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}


unusual_days <- wmata_zscores %>%
  filter(abs(rail_zscore) > 2) %>%
  arrange(desc(abs(rail_zscore)))

unusual_days %>%
  select(date, rail, rail_zscore) %>%
  kable(digits = 2)

unusual_days |>
  summarise(
    count = n()
  )
```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(1)
sample91 = wmata |> sample_n(91) 

sample91 |>
    summarise(
    bussample_mean = mean(bus),
    railsample_mean = mean(rail),
    bussample_sd = sd(bus),
    railsample_sd = sd(rail)
  )



```
#The stats are relatively close with the means only being about 10000 off, which is not a ton in this data. The standard deviation of the rail was almost spot on but the standard deviation of the bus was 3000 off, which is still not a ton. This suggests that even though we had a huge population, our sample was big enough to be close to accurate, but not big enough to be exactly accurate. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)




## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")

```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |>
  mutate(
    theft_rate2023 = thefts_2023/population * 10000, theft_rate2022  = thefts_2022/population * 10000
  )
```
A per campita rate is more useful beacuse than raw counts because larger counties will have seemingly inflated numbers, whereas smaller counties with a higher crime rate might fly under the radar. 

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts|>
  summarise(
   median_2023 = median(thefts_2023),
   total2023 = sum(thefts_2023),
   median2022 = median(theft_rate2022),
   total2022 = sum(thefts_2022)
  )
 md_car_thefts |>
  group_by(
    filter(theft_rate2023 > 105.5)
  ) 
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)
I would prioritize using 

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)
You want to use z-scores because then you can see how many days are under performing or over performing to a degree that is unusual. If there are a lot of observations where the z-score is 2 standard deviations under the mean, than those pools must be abnormally under performing. Raw attendance can be less useful because they don't put the data into perspective of the rest of the data. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)


---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
