---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
city_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv") 

```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
city_crime_rates <- city_crime |>
mutate(
violent_rate_2019 = ((violent2019 / pop2019) * 100000) , 
property_rate_2019 = ((property2019 / pop2019) * 100000) ,
violent_rate_2020 = ((violent2020 / pop2020) * 100000) , 
property_rate_2020 = ((property2020 / pop2020) * 100000)
  
)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Despite having the highest crime rate in the state, Baltimore City has seen some of the biggest drops in crime where other cities have seen steady numbers or even increases. I think the story there is how has Baltimore lowered crime rates, is there really less crime in the city? 

Frederick is the second largest city by population in the data and has the highest cuts in crime in both categories. The story here would be how they managed to do that? Because of policing or COVID?


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- city_crime_rates |>
mutate(
  violent_pct_change = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) * 100  ,
  property_pct_change = ((property_rate_2020 - property_rate_2019) / property_rate_2019) * 100 
)

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

# Despite the pandemic, violent crime increased in many places and property crime decreased in many places. Even among the largest cities, property crime was down in most places. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

# In a time when people are fighting for toilet paper and survival, property crime in most Maryland cities dropped from 2019 to 2020. 
#Over the last year, property crime in Maryland dropped in 70% of cities across the state. Frederick leads the major cities with a 95% drop in violent crime, and a 90% drop in property crimes. Even Baltimore, a city that has struggled with crime in the past, saw a 12% decrease in violent crime, and a 25% decrease in property crime. Could these decreases be fortunate side effects of a deadly pandemic, or is there something else at play? 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
